---
title: "Assesment"
output: html_document
date: "2024-11-01"
---
# Pregnancy termination trend over 6 year period in light of contraceptives dispensation trends

Birth control availability was a liberating step towards women having full control of their reproduction. However, in recent years there has been a decreasing trend in oral contraception usage [reference]. Some of these changes might be attributed to social media discussions on these topics [https://www.sciencedirect.com/science/article/pii/S0277953623004380#bib15]. Extensive list of side effects have been brought into light and dissmissive attitude of general health practitioners when it comes to a contraceptive choice has been discussed. On the one hand, it empowers women with knowledge to make an informed choice, while on the other hand it creates a space for missinformation and skewed experiences getting more representation than the general ones. 

Nevertheless, female contraceptives remain being the most reliable way to prevent unwanted pregnancy [reference]. With its effects being brought into focus it probes a question of how the trends of contraceptives usage relate to pregnancy termination statistics. 

This report will explore contraceptive trends used by females in Scotland between 2018 and 2023 and look how this trend correlated with pregnancy termination statistics.


### AIMS:

1. To see if there is a national trend in short-term (oral, injections, and implants) and long-term (IUD and coil) contraceptives usage in women of reproductive age in Scotland;
2. To see if a similar trend is maintained in different health boards;
3. Look at the correlation between contraceptive trend and pregnancy termination trend nation wide and within health boards.
4. Discuss the possible 


Hypothesis: contraceptive usage decline in some way can be associated with increase in pregnancy termination.

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
```


```{r Used packages, include=FALSE}
library(tidyverse)
library(janitor) # cleaning data
library(gt) # tables
library(here) # directory structure (will be useful later)
library(vroom)
library(stringr)
library(viridis)
```

## Data used in the report

Abortion and prescription data used in this report has been obtained from [Public Health Scotland](https://www.opendata.nhs.scot/dataset). For showing the contraceptive usage trend in 6 years all age data was used that includes females under 20 to females over 40. 

To ensure comparable figures between different boards the number of pregnancy termination events was normalised per 1000 of women in the health board in accordance to 2022 Cencus data. Only women of biological reproductive age were included as they are the most likely to be included in the statistic. The chosen respoductive age span was from 13 (average age of first menstruation) to 50 (average age of natural menopause) [https://academic.oup.com/hropen/article/2022/2/hoac005/6527206#356986178].



```{r Heatlh board names and population, include=FALSE}
health_boards <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names() %>% 
  select(hb, hb_name)

health_board_population <- read_csv(here( "data", "UV103_age_health_board_census.csv"), skip = 10) %>% 
  clean_names() %>% 
  rename(hb_name = "health_board_area_2019") %>% 
  filter(age %in% c(13:50), sex == "Female") %>% 
  group_by(hb_name) %>% 
  summarise(female_repro_population = sum(count)) %>% 
  mutate(hb_name = paste("NHS", hb_name))
```



```{r Abortion data, include=FALSE}
#data on abortion 2000-2023
abortion_data <- read_csv("https://www.opendata.nhs.scot/dataset/d684d4a5-f7ae-4a1a-ae8d-adf55304274e/resource/342f9627-dfdd-41f5-a27c-0a3c7bcb8672/download/residence_age_2023.csv") %>% 
  clean_names() %>% 
  rename(hb = "hbr", year = "yearof_termination") %>% 
  filter(age_group == "All Ages") %>% 
  select(year, hb, numberof_terminations)

abortion_data <- left_join(health_boards,abortion_data,  by = "hb") 
abortion_data <- left_join(abortion_data, health_board_population, by = "hb_name") 

abortion_data <- abortion_data %>% 
  mutate(abortion_per_1000_women = numberof_terminations/female_repro_population * 1000)


```

Prescription data has been filtered based on BNF item code. Codes beginning with 0703 and 21040 were selected as representing contraceptive medication and devices. 

As for pregnancy termination data, number of items count was normalised per 1000 of women of reproductive age in the health board based on 2022 Cencus data. 

```{r URLs for prescription data, include=FALSE}
#prescription data between 2018 and 2023 in a names vector

prescription_data_urls <- c(
# List for 2023
  "dec_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/00cdf8d7-f784-4e87-894c-34f2540ea6ab/download/pitc202312.csv",
  "nov_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/21d11fed-a494-4e30-9bb6-46143c3b9530/download/pitc202311.csv",
  "oct_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/94134438-db1d-4b8d-8f70-5e9b0e47bd03/download/pitc202310.csv",
  "sep_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/2d3d240f-1467-4a91-9f0e-769745650cb9/download/pitc202309.csv",
  "aug_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/72ca4ad2-0228-4672-9eb0-cc911e4a8ca7/download/pitc202308.csv",
  "jul_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/7bb45ee6-6f1c-45f4-933b-958bdbe0ca4f/download/pitc202307.csv",
  "jun_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/66e9611a-fbea-45b2-9edd-351c388fd06d/download/pitc202306.csv",
  "may_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a0cf1a9f-d8c3-4bda-a8d9-733897c4c288/download/pitc202305.csv",
  "apr_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/720699b0-7584-4ddb-9915-79b298189d1d/download/pitc202304.csv",
  "mar_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/8dd06c58-1a09-483a-8a01-5d68cfb8b38e/download/pitc202303.csv",
  "feb_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/ad6d7f4e-08fd-40ed-ad49-bf450d386f39/download/pitc202302.csv",
  "jan_2023" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/6caa7144-f0e4-4ab0-b9e4-8fa10c8edf1c/download/pitc202301.csv",

# List for 2022
  "jan_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/53a53d61-3b3b-4a12-888b-a788ce13db9c/download/pitc202201.csv",
  "feb_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/bd7aa5c9-d708-4d0b-9b28-a9d822c84e34/download/pitc202202.csv",
  "mar_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a0ec3bf2-7339-413b-9c66-2891cfd7919f/download/pitc202203.csv",
  "apr_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/7de8c908-86f8-45ac-b6a4-e21d1df30584/download/pitc202204.csv",
  "may_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/1b4e3200-b6e6-415f-b19a-b9ef927db1ab/download/pitc202205.csv",
  "jun_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/debeadd8-2bbb-4dd3-82de-831531bab2cb/download/pitc202206.csv",
  "jul_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/26ce66f1-e7f2-4c71-9995-5dc65f76ecfb/download/pitc202207.csv",
  "aug_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/49fa5784-be06-4015-bc6d-9b5db8726473/download/pitc202208.csv",
  "sep_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9d0a518d-9d9c-4bcb-afd8-51f6abb7edf1/download/pitc202209.csv",
  "oct_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/bd7bc2cf-4de5-4711-bd5a-9e3b77305453/download/pitc202210.csv",
  "nov_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/023986c0-3bb2-43cb-84e8-2e0b3bb1f55f/download/pitc202211.csv",
  "dec_2022" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/00213ffa-941e-4389-9e6f-3bca8067da8c/download/pitc202212.csv",

# List for 2021
  "jan_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/7722a6e6-a6b6-49ec-aa63-fdc4bc727f05/download/pitc202101.csv",
  "feb_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/3f5c55ac-1bcd-4c57-a7b6-12911f15239c/download/pitc202102.csv",
  "mar_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/df6fc708-5c50-4d57-a5c4-faa19a92c227/download/pitc202103.csv",
  "apr_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/51b7ad3f-6d52-4165-94f4-92e322656c85/download/pitc202104.csv",
  "may_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/167ffab7-a168-43d1-90c4-118aa955edfb/download/pitc202105.csv",
  "jun_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/05159d26-f504-47ce-9d33-29b739e666ea/download/pitc202106.csv",
  "jul_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/5c1cd207-1958-4105-923f-ab60983d3f90/download/pitc202107.csv",
  "aug_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/6cdae245-0423-4e6f-9e1c-dc9e129df3aa/download/pitc202108.csv",
  "sep_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/d7d1ada5-2763-4698-bf39-7bdb06f67377/download/pitc202109.csv",
  "oct_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/35cbc6b1-3462-4563-88ba-d57c03782534/download/pitc202110.csv",
  "nov_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/6ba23bd1-f53b-4946-bc79-00633239d08f/download/pitc202111.csv",
  "dec_2021" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/ad9e7b46-47fb-4d42-baad-f8e98e8f5936/download/pitc202112.csv",

# List for 2020
  "jan_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e5c841f2-3e16-428b-97db-0798ec7a5fb4/download/pitc202001.csv",
  "feb_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/af121e7c-4124-4955-92e9-a9580ccd469e/download/pitc202002.csv",
  "mar_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9581bd8d-5568-4462-93a6-6d7bfbbe7cbc/download/pitc202003.csv",
  "apr_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9cdc0526-21ab-43de-b832-bc032cd31b24/download/pitc202004.csv",
  "may_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f2ffa7b3-3f93-470a-8125-880afb9aafe0/download/pitc202005.csv",
  "jun_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/64131d22-ab47-43ff-9674-ebe8ed7edbd7/download/pitc202006.csv",
  "jul_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/80b6ddb1-f09b-4d76-9927-da1e118b01ff/download/pitc202007.csv",
  "aug_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/5fd5f56e-d1d2-4f22-b4f7-224fbd50dca3/download/pitc202008.csv",
  "sep_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e1b0cfdd-d184-4ebc-bd93-f1a10d84ead0/download/pitc202009.csv",
  "oct_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/83bf4892-d246-41f8-a6ca-d44d18e2a2ca/download/pitc202010.csv",
  "nov_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/ab8aa642-2562-4abb-a360-5c5f9e7e349c/download/pitc202011.csv",
  "dec_2020" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/0c033702-4d88-4f2d-989c-a709b1f4529e/download/pitc202012.csv",

# List for 2019
  "jan_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/3bd6e3cc-b8b7-493b-b6aa-f9fcadbb05d2/download/pitc201901.csv",
  "feb_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/3729a5c8-787a-46e3-ab7d-feed9aa91b4e/download/pitc201902.csv",
  "mar_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/87b21840-beb8-40fd-bad8-579643bc8b8b/download/pitc201903.csv",
  "apr_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/02197246-5d98-4ba9-b25d-218ac9cd91e6/download/pitc201904.csv",
  "may_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/7479536b-0b95-43d6-9152-31bbd522e6b4/download/pitc201905.csv",
  "jun_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/6ea2f299-76bc-49cd-ab43-9228b601da5f/download/pitc201906.csv",
  "jul_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/6e3856e9-88cb-495a-8c8a-54b0460df950/download/pitc201907.csv",
  "aug_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/5c667230-4201-4a09-949d-3e6cc3a4ec19/download/pitc201908.csv",
  "sep_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/ab3c15a9-44da-4a85-bb7a-58a8ce11f3d8/download/pitc201909.csv",
  "oct_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/55d93898-5382-4a59-a42e-bbc68f48b217/download/pitc201910.csv",
  "nov_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/eeee4a8f-3439-425a-aacd-58e536da90a0/download/pitc201911.csv",
  "dec_2019" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/fa276ad2-669a-472f-9c47-809f199fae21/download/pitc201912.csv",

# List for 2018
  "jan_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/82ef2c55-5a31-4759-ab9a-83b176e107f2/download/pitc201801.csv",
  "feb_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/1ae16a5f-aa26-4911-989b-ba4a6ab87037/download/pitc201802.csv",
  "mar_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/0b7bc1dd-9731-405b-a55e-656a8c996a58/download/pitc201803.csv",
  "apr_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e35aebd7-55be-4c8b-82b7-54d1ed946b91/download/pitc201804.csv",
  "may_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a2a3d0fd-d9c4-48e9-a569-c5d8abdf76f0/download/pitc201805.csv",
  "jun_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/d73cfc66-3330-416f-8507-873db28eca5c/download/pitc201806.csv",
  "jul_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/fec5dfdb-96dc-4e87-b7a0-6ecf20ec86e9/download/pitc201807.csv",
  "aug_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/592f3b58-2da8-4cc7-b4ab-e0eb0f16aa3f/download/pitc201808.csv",
  "sep_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/bd18b099-942d-42fb-9fa5-b947714381cd/download/pitc201809.csv",
  "oct_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/37e94a73-121e-443e-9564-2e1d9e36c5eb/download/pitc201810.csv",
  "nov_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9cc56231-f184-4e29-a5ef-e6080bc78e93/download/pitc201811.csv",
  "dec_2018" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/c50ed972-34e8-4cd6-8d2e-349e875c3ffa/download/pitc201812.csv",
 
# List for 2017 
  "jan_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/3def845e-c830-4b73-8e02-e42adcde5afa/download/pitc201701.csv",
  "feb_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f7a00faf-49ec-4a99-9686-e52b2d1572c4/download/pitc201702.csv",
  "mar_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/06c958d6-bae7-4d1b-9b62-88db123dc7b6/download/pitc201703.csv",
  "apr_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/78ff3279-2486-497c-968c-256a12631088/download/pitc201704.csv",
  "may_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/014bf682-7413-4418-abc6-233e145dd454/download/pitc201705.csv",
  "jun_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/13042cf7-fa4e-464c-aa9f-18c5213fb7b0/download/pitc201706.csv",
  "jul_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/44bc24dd-3a8e-4056-8ee2-eb903bc352ac/download/pitc201707.csv",
  "aug_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/ff0da464-d78f-4394-add2-101e78820932/download/pitc201708.csv",
  "sep_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/106c27b7-4615-4a14-a7d9-8dc22f86dbd2/download/pitc201709.csv",
  "oct_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/01cdd69a-049a-46cf-a618-75b5f93422f1/download/pitc201710.csv",
  "nov_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/0d9b6b7d-bd94-4d72-8806-22025edcdeeb/download/pitc201711.csv",
  "dec_2017" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a39b254a-ab6a-41d5-a920-5ff34bdb9f6b/download/pitc201712.csv",

# List for 2016
  "jan_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/7c487db9-f450-4319-8b18-4b825380692b/download/pitc201601.csv",
  "feb_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/0727befa-5530-4c1d-aff0-ccb7554eff7a/download/pitc201602.csv",
  "mar_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/20ad9575-4bd7-4bf4-af41-fd36071e2301/download/pitc201603.csv",
  "apr_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e44eb789-b8e0-41d9-a264-eba18b2d2473/download/pitc201604.csv",
  "may_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/90ebb5ed-bc22-4bbc-8915-d55ae9d9906e/download/pitc201605.csv",
  "jun_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a636862a-77e0-4c97-ba97-268578534a8e/download/pitc201606.csv",
  "jul_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/c12c905b-3a2d-4b7a-b5d2-8c359b786beb/download/pitc201607.csv",
  "aug_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/bbd1d028-b9c6-4a28-b579-eafcaaa44703/download/pitc201608.csv",
  "sep_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/2e0d6f14-3752-4c42-9a02-34d867132737/download/pitc201609.csv",
  "oct_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f22c20c1-6066-4f7d-a939-3ae654d1467a/download/pitc201610.csv",
  "nov_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/5d421ba3-ccba-4d98-9772-e2e042b29a0f/download/pitc201611.csv",
  "dec_2016" = "https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/597ea4bc-f733-45b6-8a60-5d7a6ac4824c/download/pitc201612.csv"

)
```



```{r Data reading and cleaning}
columns_needed_from_prescription <- c("BNFItemCode", "NumberOfPaidItems")

read_and_clean <- function(url) {
  # vroom used as a faster function for reading delimited files
  # only columns with prescription drug name, health board, and number of items paid selected to make loading onto R more memory efficient
    #since older data files have health board column named HBT2014 while newer have is named HBT, any_of() is used to select only the ones that exist in the file being read
  data <- vroom(url, col_select = any_of(c(columns_needed_from_prescription, "HBT", "HBT2014")), skip_empty_rows = TRUE)
  
 
  #data filtered for contraceptives (including contraceptive devices) only (based on the BNF code)
  data_filtered <- data %>%
  filter(str_starts(BNFItemCode, "0703") | str_starts(BNFItemCode, "21040")) %>% 
    clean_names()
  colnames(data_filtered)[3] ="hb"

  
  # assign function used to create a value with the name of the url from named list and assign read preprocessed data to that variable
name <- names(prescription_data_urls)[which(prescription_data_urls == url)] 
assign(name, data_filtered, envir = .GlobalEnv) 
}

lapply(prescription_data_urls, read_and_clean)
```

```{r Combine all prescription datases}
#sum the number of items paid for by health board
add_year <- function(url) {
  name <- names(prescription_data_urls)[which(prescription_data_urls == url)] 
  data <- get(name) 
  data <- data %>% 
    add_column(year = as.numeric(str_extract_all(name, "\\d+")))
  
  
  object_name <- paste(name, "sum", sep = "_")
  assign(object_name, data, envir = .GlobalEnv) 
}


lapply(prescription_data_urls, add_year)

#join full prescription dataset with healh board names and populations of women of reproductive age
full_prescirption_dataset <- bind_rows(mget(paste(names(prescription_data_urls), "sum", sep = "_")))
full_prescirption_dataset <- left_join(health_boards, full_prescirption_dataset, by = "hb")
full_prescirption_dataset <- left_join(full_prescirption_dataset, health_board_population, by = "hb_name")
```

The data was then further devided into long term contraceptives (IUD, coil, and implant) lasting several years and short term contraceptives that have to be taken regularly (oral tablets, cervical caps, and injections). The division was based on the BNF codes with 0703022P representing implants, 0703023 representing IUDs, and 210400002 representing copper coils. The remaining birth control was assigned to short-term cathegory. Same as abortion data the contraceptive data was normalised by women of reproductive age number in the health board for health board comparison.

```{r filter datasets to create long term and short term contraception datasets}
long_term_contraception <- full_prescirption_dataset %>% 
  filter(str_detect(bnf_item_code, "0703022P|0703023|210400002")) %>% 
  group_by(year, hb_name, female_repro_population) %>% 
  summarise(sum_of_contraceptive_items = sum(number_of_paid_items, na.rm = TRUE)) %>% 
  mutate(contracep_items_per_1000_women = sum_of_contraceptive_items/female_repro_population*1000)

short_term_contraception <- full_prescirption_dataset %>% 
  filter(!str_detect(bnf_item_code, "0703022P|0703023|210400002")) %>% 
  group_by(year, hb_name, female_repro_population) %>% 
  summarise(sum_of_contraceptive_items = sum(number_of_paid_items, na.rm = TRUE)) %>% 
  mutate(contracep_items_per_1000_women = sum_of_contraceptive_items/female_repro_population*1000)

```


## Results

### Contraceptives trends

It has been shown that over the past 8 years there has been a decline in combined oral contraceptives being prescribed while the progesterone only pill prescriptions increased [https://srh.bmj.com/content/early/2024/05/24/bmjsrh-2024-202387]. While it shows that women preference might be shifting away from estrogen containing products it doesn't tell much about the overall trend in contraceptive usage. 

Figure 1 shows that there is a nationwide decline in birth control usage. There is a steep drop from pre-Covid to Covid year. It is hard to day whether the decline from year 2022 to 2023 is still due to Covid or if it represents a general trend of decline. Seeing that there was ______ over pre-Covid years it could be suspected that ______ .

```{r Figure 1 Gross contraception usage}
full_prescirption_dataset %>% 
  group_by(year) %>% 
  summarise(sum=sum(number_of_paid_items)) %>% 
  ggplot(aes(y=sum, x=year)) +
  geom_line(color="#69b3a2", size=2) +
  theme_classic() +
  theme(plot.title = element_text(size = 13, 
                                  hjust = 0,
                                  face = "bold", 
                                  margin = margin(10,10,10,10)), 
                                  axis.title.x = element_text(size = 11, margin=margin(10,0,0,0)), 
                                  axis.title.y = element_text(size = 11, margin=margin(0,10,0,0))) +
  labs(title = "Figure 1. Gross prescription of contraceptives between 2018 and 2023", x = "Year", y = "Number of items dispensed") +
  geom_vline(xintercept = 2020, color="grey", size=.5, linetype = "dashed") +
  geom_vline(xintercept = 2022, color="grey", size=.5, linetype = "dashed") +
  annotate(geom="text", x=2021, y=826000, 
             label="--------- COVID period ---------", colour = "#71797E") 
```

### Long-term and short term contraceptives

Women have multiple options when it comes to birth control, including short term options like oral pill and long-term options like IUD the coil. However, some women report feeling like they were not actually given a choice when it came to their birth control [https://www.sciencedirect.com/science/article/pii/S0277953623004380#bib15]. They report feeling like their GP's prescribed oral contraceptives by default. (With this coming to light women are taking control of their heatlh>>>>> Looging at Scotland wide trends of long term versus short term contrateption it can be seen that short-term contraception is on a consistent decline over the past ___ years. Covid doesn't seem to have had much of an impact on these prescriptions. However, when it comes to long-term contraceptives it is clear that covid hindered their use. That is likely because application of intravaginal devices and impants have to be performed by medical professional and during covid clinics' work was focused on emergency medical care. 

```{r}
long_term_contraception %>% 
  summarise(sum=sum(number_of_paid_items)) %>% 
  ggplot(aes(y=sum, x=year)) +
  labs(title = "plot_prescription_long_term") +
  geom_line()
plot_prescription_long_term

short_term_contraception %>% 
  
  ggplot(aes(y = contracep_items_per_1000_women, x = year) +
  labs(title = "plot_prescription_short_term") +
  scale_colour_manual(values = rainbow(14)) +
  geom_line() 

```



```{r Extract information from prescription datasets}

plot_prescription_long_term <- long_term_contraception %>% 
  ggplot(aes(y = contracep_items_per_1000_women, x = year, colour = hb_name)) +
  labs(title = "plot_prescription_long_term") +
  geom_line()
plot_prescription_long_term

plot_prescription_short_term <- short_term_contraception %>% 
  ggplot(aes(y = contracep_items_per_1000_women, x = year, colour = hb_name)) +
  labs(title = "plot_prescription_short_term") +
  scale_colour_manual(values = rainbow(14)) +
  geom_line() 
plot_prescription_short_term


plot_abortion <- abortion_data %>% 
  filter(year %in% c(2018:2023)) %>% 
  ggplot(aes(y = abortion_per_1000_women, x = year, colour = hb_name)) +
  scale_colour_manual(values = rainbow(11)) +
  geom_line() 
plot_abortion
```

